The Issue of Homelessness in Toronto

### 2022-02-06 outputs/paper/paper.Rmd
final edits

---
title: "The Issue of Homelessness in Toronto"
author: "Jack McKay"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::pdf_document2
toc: FALSE
subtitle: "And the changes that must be made to fix it"
abstract: "Shelter System Flow data was collected by the City of Toronto Open Portal, in order to record what populations are most affected by homelessness in Toronto as well as monitor how the City’s Shelter System is functioning. Shelter usage is an important factor in determining the livability and accessibility of a city, and monitoring trends and fluctuations can allow us to better cater the services provided to the populations of people that need it the most."
thanks: 'Code and data are available at [github.com/jmacattack27/ModellingHomelessPopulations](https://github.com/jmacattack27/ModellingHomelessPopulations).'
bibliography: references.bib
---

@@

During the 2021 elections, there was one issue that every major party unanimously agreed upon; Canada is in the midst of a severe housing crisis, and the cost of home ownership must be lowered. Since the early 2000s, Canada experienced an intense surge in housing prices. Known as the Canadian Housing Crisis, the City of Toronto has been one of the driving forces in this surge. In the last two decades, home prices in Canada have increased by 375%, while prices in Toronto have soared over 450% (@clarke). One of the driving factor of this is the increasing commodification of houses, wherein foreign investors purchase available real estate and hold it as an investment. This greatly restricts the supply of available housing in Toronto, and has a myriad of negative consequences for citizens. One such consequence is the rapidly growing homeless population in Toronto. As is commonplace with socioeconomic issues, the ones most affected by this crisis are those who are most the most economically vulnerable. Increasingly, the trend of housing prices grows more and more detached from the trends of income prices, leading housing to become less affordable to a greater number of people as this crisis progresses. The problem of Toronto's growing homeless population has reared it's ugly head at the City over the last 2 years, as Covid-19 fears and lack of adequate social distancing in shelters caused a huge outflow of people onto the streets. This made visible to all the increasing urban destitution that has been largely driven by the housing crisis. Indeed, as this problem becomes unavoidable to even the most neglecting lawmakers, the City has responded with increasingly aggressive and vacuous measures; in the summer of 2021, the city spent $2 million on clearing homeless encampments (@draaisma). Rather than funding the swift and often brutal evictment of homeless people by overpaid and overequipped police officers, the City of Toronto must instead invest in its shelter system, and improve services aimed at eliminating homelessness at its roots.

@@

The first aspect of the data I investigated relates to the success rate of permanent housing placements, ie; the likelihood that a homeless person moved to permanent housing does not return to the shelter. Figure \@ref(fig:housingattrition) graphs the number of people moved into permanent housing against the number of people who have returned to a shelter from after being moved to permanent housing. There has been a sharp decrease in the number of people being moved to housing since the beginning of 2020, while the number of people returning to shelters after being placed in permanent housing has remained relatively stable. This indicates that in the success rates of housing placements are declining, which is one of the primary gauges of how successful Toronto's homeless services are at helping citizens escape homelessness.

@@

The final aspect of the data that I analyzed was the fluctuations in the populations of various age groups of homeless people. Figure \@ref(fig:agegroups) shows how the number of homeless people in each age group has varied since January 2020. Here we can see that by far the most represented age groups are 25 to 44 and 45 to 64, which is not unexpected. As well, we notice a significant drop off in most age groups starting in March 2020, with the most drastic decreases being in people aged 15 or younger, 25 to 44 and 45 to 64. Between March 2020 and July 2020, these populations declined from 1440 to 911 (36.7%), 3893 to 2962 (23.9%) and 2852 to 2331 (18.3%), respectively. This is perfectly in line with the announcement of the first lockdown in Toronto. Interestingly, while there still was a decrease in registered people aged 16 to 24, it is much less drastic than the previously mentioned age groups, despite being in the middle of two of them. As well, we notice that there is a rebound in populations aged 25 to 44 and 45 to 64 starting around June 2020, before declining again near the start of 2021. Notably, the age group 16 and under did not experience this rebound, and had a continuous decline from March 2020 to May 2021, during which point its number fell from 1440 to 569, a colossal 60.5% decrease. However since May 2021, the population has been steadily increasing, and as of December 2021 is 901, 62.6% of its pre-lockdown total. Finally, we note that the population aged 65 or older was almost completely unaffected by Covid, holding steady between 463 and 551 January 2020 to June 2021, since which point it has increased to a 2 year high of 637 in December 2021.

@@

Initial results indicate that there are a few main areas that the city should focus its efforts on improving. The success rate of permanent housing placements has been steadily declining over the last 2 years. Finding permanent housing is the ultimate goal of 
most homeless people, and finding permanent housing, as well as helping homeless people prepare for permanent housing, should be the primary goal of homeless shelters, aside from perhaps providing adequate short term necessities. The success rate of housing placements is down from an average of 85.1% in 2020 to 70.7% in 2021, a sharp decline. As a result, the city needs to focus more of its funding and time on improving services aimed at preparing homeless people for life in permanent housing. The main factor contributing to this decline is the steady decrease in housing placements; while the amount of homeless people returning from housing has remained relatively steady, the amount of homeless people being moved to permanent housing has declined from a monthly average of of 607 between January 2020 and April 2020 to 259 between September 2021 and December 2021. While there are a multitude of factors that could influence, the most obvious one is the increasing scarcity of housing in Toronto.

@@

Another immensely consequential factor that invariably skewed every aspect of the data is the Covid-19 pandemic, which has been ongoing since March 2020. This would have had a plethora of effects, not all of which can be accounted for. The most obvious would be the health concerns and resulting social distancing measures. Homeless shelters are often crowded, and so there would naturally be a large decline in the number of homeless people registering with shelter services throughout the pandemic. This effect made itself apparent with the increasing size and frequency of outdoor homeless encampments throughout Toronto, popping up in public squares, in parking lots, and most prominently in parks. As a result, the total homeless population is invariably and drastically under-represented over the last two years. The pandemic would similarly contribute to the sharp decrease in housing placements since January 2020; in fact, in \@ref(fig:housingattrition) we can see that there is a steep drop-off in placements starting in March 2020, when the first lockdown was announced. The number of placements begins to rebound by June 2020--when restrictions first began to ease up--before experiencing another sharp decline around September 2020, when restrictions returned. Indeed, the possible effects of the pandemic on this data are myriad, and impossible to fully account for. As such, we can only try and interpret how the data may have been skewed, and take this into account when we draw our conclusions.

@@

Despite the undeniable effects the pandemic would have had on homeless populations, we can still draw some solid conclusions about what the city needs to change to improve its shelter services, as well as to work towards eliminating homelessness. The first and most obvious problem that must be addressed is the aforementioned Toronto housing crisis. While there are a number of solutions that could be implemented to address this at both a municipal and federal level, I will focus on one main avenue; preventing commodification of real estate. The commodification of real estate can be described as approaching real estate (houses, plots of land or otherwise) as an investment, as opposed to a livable property. This occurs when foreign investors, most notably from Asia, buy up properties in in areas with high demand for real estate--such as Toronto--and hold onto them as an appreciating investment. A result of this is that livable housing becomes substantially scarcer, and as a result the prices of available properties increase. Not only do these foreign investors buy up available real estate, but they commonly demolish the pre-existing, often affordable infrastructure and build more expensive (and thus profitable) rental properties. Thus, commodification not only leads to a decreased supply of purchasable homes, but also a decreased supply of affordable rental properties. Indeed, shelter is considered a fundamental human right, and therefore the commodification of real estate is the commodification and financialization of a basic human right into an investment from which to extract value. In fact, between January and August of 2021, investors accounted for more than 25% of all real estate transactions (@merali), an outrageous statistic when one considers the rapidly growing proportion of Torontonians unable to afford a home. Currently, there are no restrictions on who can buy property in Toronto, meaning that a foreign investor who has never set foot in Canada and lives full-time in another country is able to purchase property with relative ease. The first solution I propose to Toronto's housing crisis needs to happen at the provincial and/or federal level, and it involves instituting a steep tax on property purchases by foreign investors, as well as heavily taxing the profit made on these investments. This would make purchasing property as an investment in Toronto (and indeed all of Canada) far less appealing. Additionally, the federal government should alter its tax structure so as to help enable and encourage non-profit and co-op sectors to purchase more property. This would allow for an increased supply of affordable housing, as these sectors are not focused on extracting as much value as possible from Canadian citizens, but rather providing an essential and fundamental human right to those in need. Implementing these changes would not only greatly increase the market supply of housing in Toronto and throughout Canada, but also would lead to an increased supply of housing specifically designed to be affordable, two things that I believe would do wonders for homelessness in Toronto.

>>>>

---
title: "The Issue of Homelessness in Toronto"
author: "Jack McKay"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::pdf_document2
toc: FALSE
subtitle: "And the steps that can be taken to improve it"
abstract: "Shelter System Flow data was collected by the City of Toronto Open Portal, in order to record what populations are most affected by homelessness in Toronto as well as monitor how the City’s Shelter System is functioning. Shelter usage is an important factor in determining the livability and accessibility of a city, and monitoring trends and fluctuations can allow us to better cater the services provided to the populations of people that need it the most."
thanks: 'Code and data are available at [github.com/jmacattack27/ModellingHomelessPopulations](https://github.com/jmacattack27/ModellingHomelessPopulations).'
bibliography: references.bib
---

@@

During the 2021 federal elections, there was one issue that every major party unanimously agreed upon; Canada is in the midst of a severe housing crisis, and the cost of home ownership must be lowered. Since the early 2000s, Canada experienced an intense surge in housing prices. Known as the Canadian Housing Crisis, the City of Toronto has been one of the driving forces in this surge. In the last two decades, home prices in Canada have increased by 375%, while prices in Toronto have soared over 450% (@clarke). One of the driving factors of this is the increasing commodification of houses, wherein foreign investors purchase available real estate and hold it as an investment. This greatly restricts the supply of available housing in Toronto, and has a myriad of negative consequences for citizens. One such consequence is the growing homeless population in Toronto. As is commonplace with socioeconomic issues, the ones most affected by this crisis are those who are most the most economically vulnerable. Increasingly, the trend of housing prices grows more and more detached from the trends of income prices, leading housing to become less affordable to a greater number of people as this crisis progresses. The problem of Toronto's growing homeless population has reared it's ugly head at the City over the last 2 years, as Covid-19 fears and lack of adequate social distancing in shelters caused a huge outflow of people onto the streets. This made visible to all the increasing urban destitution that has been largely driven by the housing crisis. Indeed, as this problem becomes unavoidable to even the most neglecting lawmakers, the City has responded with increasingly aggressive and ineffective measures; in the summer of 2021, the city spent $2 million on clearing homeless encampments (@draaisma). Rather than funding the swift and often brutal evictment of homeless people by overequipped police officers, the City of Toronto must instead invest in its shelter system, and improve services aimed at eliminating homelessness at its roots.

@@

The first aspect of the data I investigated relates to the success rate of permanent housing placements, ie; the likelihood that a homeless person moved to permanent housing does not return to the shelter. Figure \@ref(fig:housingattrition) graphs the number of people moved into permanent housing against the number of people who have returned to a shelter after being moved to permanent housing. There has been a sharp decrease in the number of people being moved to housing since the beginning of 2020, while the number of people returning to shelters after being placed in permanent housing has remained relatively stable. This indicates that in the success rates of housing placements are declining, which is one of the primary gauges of how successful Toronto's homeless services are at helping citizens escape homelessness.

@@

The final aspect of the data that I analyzed was the fluctuations in the populations of various age groups of homeless people. Figure \@ref(fig:agegroups) shows how the number of homeless people in each age group has varied since January 2020. Here we can see that by far the most represented age groups are 25 to 44 and 45 to 64, which is not unexpected. As well, we notice a significant drop off in most age groups starting in March 2020, with the most drastic decreases being in people aged 15 or younger, 25 to 44 and 45 to 64. Between March 2020 and July 2020, these populations declined from 1440 to 911 (36.7%), 3893 to 2962 (23.9%) and 2852 to 2331 (18.3%), respectively. This is perfectly in line with the announcement of the first lockdown in Toronto. Interestingly, while there still was a decrease in registered people aged 16 to 24, it is much less drastic than the previously mentioned age groups, despite being in the middle of two of them. As well, we notice that there is a rebound in populations aged 25 to 44 and 45 to 64 starting around June 2020, before declining again near the start of 2021. Notably, the age group 16 and under did not experience this rebound, and had a continuous decline from March 2020 to May 2021, during which time its number fell from 1440 to 569, a colossal 60.5% decrease. However since May 2021, the population has been steadily increasing, and as of December 2021 is 901, 62.6% of its pre-lockdown total. Finally, we note that the population aged 65 or older was almost completely unaffected by Covid, holding steady between 463 and 551 January 2020 to June 2021, since which point it has increased to a 2 year high of 637 in December 2021.

@@

Initial results indicate that there are a few main areas that the city should focus its efforts on improving. The success rate of permanent housing placements has been steadily declining over the last 2 years. Finding permanent housing is the ultimate goal of 
most homeless people, and finding permanent housing, as well as helping homeless people prepare for permanent housing, should be the primary goal of homeless shelters, aside from perhaps providing adequate short term necessities. The success rate of housing placements is down from an average of 85.1% in 2020 to 70.7% in 2021, a sharp decline. As a result, the city needs to focus more of its funding and time on improving services aimed at preparing homeless people for life in permanent housing. The main factor contributing to this decline is the steady decrease in housing placements; while the amount of homeless people returning from housing has remained relatively steady, the amount of homeless people being moved to permanent housing has declined from a monthly average of of 607 between January 2020 and April 2020 to 259 between September 2021 and December 2021. While there are a multitude of factors that could influence, the most obvious one is the increasing scarcity of affordable housing in Toronto.

@@

Another immensely consequential factor that invariably skewed every aspect of the data is the Covid-19 pandemic, which has been ongoing since March 2020. This would have had a plethora of effects, not all of which can be accounted for. The most obvious would be the health concerns and resulting social distancing measures. Homeless shelters are often crowded, and so there would naturally be a large decline in the number of homeless people registering with shelter services throughout the pandemic. This effect made itself apparent with the increasing size and frequency of outdoor homeless encampments throughout Toronto, popping up in public squares, in parking lots, and most prominently in parks. As a result, the total homeless population is invariably and drastically under-represented over the last two years. The pandemic would similarly contribute to the sharp decrease in housing placements since January 2020; in fact, in Figure \@ref(fig:housingattrition) we can see that there is a steep drop-off in placements starting in March 2020, when the first lockdown was announced. The number of placements begins to rebound by June 2020--when restrictions first began to ease up--before experiencing another sharp decline around September 2020, when restrictions returned. Indeed, the possible effects of the pandemic on this data are myriad, and impossible to fully account for. As such, we can only try and interpret how the data may have been skewed, and take this into account when we draw our conclusions.

@@

Despite the undeniable effects the pandemic would have had on homeless populations, we can still draw some solid conclusions about what the city needs to change to improve its shelter services, as well as to work towards eliminating homelessness. The first and most obvious problem that must be addressed is the aforementioned Toronto housing crisis. While there are a number of solutions that could be implemented to address this at both a municipal and federal level, I will focus on one main avenue; preventing commodification of real estate. The commodification of real estate can be described as approaching real estate (houses, plots of land or otherwise) as an investment, as opposed to a livable property. This occurs when foreign investors, buy up properties in in areas with high demand for real estate--such as Toronto--and hold onto them as an appreciating investment. A result of this is that livable housing becomes substantially scarcer, and as a result the prices of available properties increase. Not only do these foreign investors buy up available real estate, but they commonly demolish the pre-existing, often affordable infrastructure and build more expensive (and thus profitable) rental properties. Thus, commodification not only leads to a decreased supply of purchasable homes, but also a decreased supply of affordable rental properties. Indeed, shelter is considered a fundamental human right, and therefore the commodification of real estate is the commodification and financialization of a basic human right into an investment from which to extract value. In fact, between January and August of 2021, investors accounted for more than 25% of all real estate transactions (@merali), an outrageous statistic when one considers the rapidly growing proportion of Torontonians unable to afford a home. Currently, there are no restrictions on who can buy property in Toronto, meaning that a foreign investor who has never set foot in Canada and lives full-time in another country is able to purchase property with relative ease. The first solution I propose to Toronto's housing crisis needs to happen at the provincial and/or federal level, and it involves instituting a steep tax on property purchases by foreign investors, as well as heavily taxing the profit made on these investments. This would make purchasing property as an investment in Toronto (and indeed all of Canada) far less appealing. British Columbia currently has a 20% foreign buyer tax on all profits earned on a real estate investment (@govbc), and I believe that Ontario should implement something similar. In addition to slowing down foreign investment, the money made from this tax could then go towards funding government projects aimed at building affordable housing projects.

The federal government should alter its tax structure so as to help enable and encourage non-profit and co-op sectors to purchase more property. This would allow for an increased supply of affordable housing, as these sectors are not focused on extracting as much value as possible from Canadian citizens, but rather providing an essential and fundamental human right to those in need. Implementing these changes would not only greatly increase the market supply of housing in Toronto and throughout Canada, but also would lead to an increased supply of housing specifically designed to be affordable, two things that I believe would be highly effective in managing homelessness in Toronto.
